Comparison of approaches to using machine learning models to increase the efficiency of the global search algorithm for solving multicriterial problems

### 2024-08-03 2024/OPTIMA/OPTIMA_2024_MCO.tex

%
\begin{abstract}
Solving a multicriteria optimization problem requires finding a whole set of independent variables corresponding to non-dominated criteria values (Pareto set). The complexity of these problems increases significantly in the case when the criteria are multiextremal. The paper presents several variations of the global optimization algorithm intended for solving ``black-box'' multiobjective problems with multiextremal criteria. The proposed methods are based on a combination of the information-statistical approach and machine learning procedures aimed at increasing the efficiency of constructing the Pareto set. As a novelty, the paper describes a new approach to using machine learning models within the algorithm of multicriteria optimization. The effectiveness of different versions of the proposed algorithm is estimated on the base of a representative computational experiment.

\keywords{Multicriteria Problems \and Global Optimization \and Machine Learning.}
\end{abstract}
%
%
%
\section{Introduction}
\label{sec:1}

@@

\begin{equation}
\label{eq:01}
  \Phi(y) = (f_1 (y),f_2 (y), \dots, f_s(y))
\end{equation}
where scalar functions $f_i(y)$, $1 \leq i \leq s$, are considered as the partial efficiency criteria, $y=(y_1,y_2, \dots ,y_N)$ is the vector of independent parameters belonging to the feasible domain $D \in R^N$ that is the hyperparallelepiped
\begin{equation}
\label{eq:02}
    D=\{y \in R^N : a_i \leq y_i \leq b_i, 1 \leq i \leq N\}
\end{equation}
in the $N$-dimensional Euclidean space, and $a_i$, $b_i$, $1 \leq i \leq N$, are constant values.
The best value of the partial criterion $f_i(y)$, $1 \leq i \leq s$, is supposed to correspond to the global minimum of this criterion in the domain $D$.
Let's write this problem in the form
\begin{equation}
\label{eq:02a}
  \Phi(y) = (f_1 (y),f_2 (y), \dots, f_s(y)) \to \min, \; y \in D.
\end{equation}

@@

There are various ways to construct the function $F(\lambda, y)$ \cite{Miettinen1999,Ehrgott2005,GergelKozinov2020,Marler2004}. In our study we utilized the maximum convolution 
\begin{equation}
\label{eq:05}
F(\lambda, y) = \max_{1 \leq i \leq s} \left\{\lambda_i f_i (y)\right\}, y \in D, \sum_{i=1}^s {\lambda_i} = 1, 0 \leq \lambda_i \leq 1.
\end{equation}

@@

\textbf{Dimensionality reduction.} For solving the optimization problems (\ref{eq:04}) the dimensionality reduction \cite{Gergel2019_2,Gergel2018,GergelKozinov2020} on the base of Peano curves is applied. The Peano curve $y(x)$ maps the close interval $[0,1]$ of the real axis onto the parallelepiped (\ref{eq:02}) unambiguously and continuously. It enables to replace the multidimensional problem (\ref{eq:04}) with solving the equivalent univariate problem
\begin{equation}
\label{eq:07}
\min \left\{\varphi(y), y \in D\right\} = \min\left\{ \varphi(y(x)), x \in [0,1] \right\}.
\end{equation}

@@

As part of the proposed approach, the machine learning model is built according to the rules similar to the construction of the SVC model from section \ref{ssec:41}. As before, all trial points are divided into two classes -- points of the Pareto set and points that do not belong to the Pareto set. The classifier is trained on the selected classes. Note that the classifier models of most machine learning libraries can return not only the class label, but also the probability of relevance to the class. Unlike the approach described in Subsection \ref{ssec:41} the characteristic $R_{ps}(i)$ is computed as the average of probabilities of situations when the points
%$y_i=y(x_i)$ and $y_{i-1}=y(x_{i-1})$
correspond to boudaries of the interval $(x_{i-1},x_{i})$ belong to the Pareto set.

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{fig1.png}
\caption{Heatmap of the function $R_{ps}(i)$ in the criteria space for algorithms with different machine learning models.} 
\label{fig:1}
\end{figure}

@@

First, when changing the optimization problem from set (\ref{eq:07}), a new weight calculation scheme for each of two classes of trial points is proposed. In this scheme, the weights of each class are calculated inversely proportional to the ratio of the number of all points to the number of points belonging to the Pareto set.

@@

Computational experiments were conducted for each method and each model in order to assess the effectiveness of using these techniques in the global search algorithm. Fig. \ref{fig:2} demonstrates the visualization of $R_{ps}(i)$ similar to Fig. \ref{fig:1}.
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{fig2.png}
\caption{Heatmap of the function $R_{ps}(i)$ in the criteria space for algorithms with different techniques of building models after transformation (\ref{eq:18}).} 
\label{fig:2}
\end{figure}

@@

\textbf{Experiments with fixed weights of two trial classes.} The first series of experiments is aimed at comparison the algorithm ML\_MGSA\_Prob (Fixed weights) with results of running for methods MGSA and ML\_MGSA\_Dist.

@@

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
% \usepackage{graphicx}
\begin{table}[ht]
\centering
\caption{Results of computational experiments}
\label{tab:1}
\resizebox{\textwidth}{!}{%
\begin{tabular}{cccccc}
\hline
\textbf{}                                                                  & \textbf{Model Kernel} & \textbf{$\alpha$} & \textbf{Avg. HV-index} & \textbf{Avg. iter.} & \textbf{Speedup} \\ \hline
ML\_MGSA                                                                   & -          & -    & 114.91  & 2269.82   & 1.00           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of fixed weights}}                                                                \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 114.67  & 691.14    & \textbf{3.28}  \\
\multirow{3}{*}{ML\_MGSA\_Prob (Fixed weights)}                            & Linear     & 0.04 & 114.72  & 963.05    & 2.36           \\
                                                                           & Poly       & 0.07 & 114.73  & 871.82    & 2.60           \\
                                                                           & RBF        & 0.03 & 114.80  & 883.04    & 2.57           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of adjusted weights}}                                                             \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 114.67  & 691.14    & \textbf{3.28}  \\
\multirow{3}{*}{ML\_MGSA\_Prob (Adj. weights)}                             & Linear     & 0.04 & 114.76  & 1018.03    & 2.23           \\
                                                                           & Poly       & 0.08 & 114.73  & 844.66    & 2.69           \\
                                                                           & RBF        & 0.03 & 114.76  & 796.74    & 2.85           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of adjusted weights and logarithmic probabilities}}                                               \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 114.67  & 691.14    & \textbf{3.28}  \\
\multirow{3}{*}{ML\_MGSA\_Prob (Log Norm)}                                 & Linear     & 0.05 & 114.72  & 711.44    & 3.19           \\
                                                                           & Poly       & 0.04 & 114.79  & 854.82    & 2.66           \\
                                                                           & RBF        & 0.04 & 114.74  & 704.99    & 3.22           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments with 4-criterial MCO problems}}                                                                \\ \hline
ML\_MGSA                                                                   & -          & -    & 7819.86 & 6046.4    & 1.00           \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 7751.43 & 1925.6    & 3.14           \\
\multirow{3}{*}{ML\_MGSA\_Prob (Log Norm)}                                 & Linear     & 0.08 & 7729.71 & 1760.5    & 3.43           \\
                                                                           & Poly       & 0.08 & 7796.20 & 4121.7    & 1.47           \\
                                                                           & RBF        & 0.08 & 7786.48 & 1364.6    & \textbf{4.43}  \\ \hline
\end{tabular}%
}
\end{table}

Fig. \ref{fig:3} shows the change of the average HV-index as a function of the number of trials.
\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{fig3.png}
\caption{Comparison of HV-indices for the case of fixed weights.} 
\label{fig:3}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{fig4.png}
\caption{HV-indices comparison  in the case of adjusted weights} 
\label{fig:4}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{fig5.png}
\caption{Comparison of  HV-indices for methods with probabilities (\ref{eq:18})} 
\label{fig:5}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=0.8\textwidth]{Fig6.png}
\caption{Efficiency comparison in the 4-critarial case} 
\label{fig:6}
\end{figure}

>>>>

%
\begin{abstract}
Solving a multicriteria optimization problem requires finding a whole set of independent variables corresponding to non-dominated criteria values (Pareto set). The complexity of these problems increases significantly in the case when the criteria are multiextremal. The paper presents several variations of the global optimization algorithm intended for approximate solving ``black-box'' multiobjective problems with multiextremal criteria. The proposed methods are based on a combination of the information-statistical approach and machine learning procedures aimed at increasing the efficiency of constructing the Pareto set approximation. As a novelty, the paper describes a new approach to using machine learning models within the algorithm of multicriteria optimization. The effectiveness of different versions of the proposed algorithm is estimated on the base of a representative computational experiment.

\keywords{Multicriteria Problems \and Global Optimization \and Machine Learning \and Peano Curves.}
\end{abstract}
%
%
%
\section{Introduction}
\label{sec:1}

@@

\begin{equation}
\label{eq:01}
  \Phi(y) = (f_1 (y),f_2 (y), \dots, f_s(y))
\end{equation}
where scalar functions $f_i(y)$, $1 \leq i \leq s$, are considered as the partial efficiency criteria, $y=(y_1,y_2, \dots ,y_N)$ is the vector of independent parameters belonging to the feasible domain $D \subset R^N$ that is the hyperparallelepiped
\begin{equation}
\label{eq:02}
    D=\{y \in R^N : a_i \leq y_i \leq b_i, 1 \leq i \leq N\}
\end{equation}
in the $N$-dimensional Euclidean space, and $a_i$, $b_i$, $1 \leq i \leq N$, are constant values.
The best value of the partial criterion $f_i(y)$, $1 \leq i \leq s$, is supposed to correspond to the global minimum of this criterion in the domain $D$.
Let's write this problem in the form
\begin{equation}
\label{eq:02a}
  \Phi(y) = (f_1 (y),f_2 (y), \dots, f_s(y)) \to \min, \; y \in D.
\end{equation}

@@

There are various ways to construct the function $F(\lambda, y)$ \cite{Miettinen1999,Ehrgott2005,GergelKozinov2020,Marler2004}. In our study we utilized the maximum convolution 
\begin{equation}
\label{eq:05}
F(\lambda, y) = \max_{1 \leq i \leq s} \left\{\lambda_i f_i (y)\right\}, y \in D, \sum_{i=1}^s {\lambda_i} = 1, 0 < \lambda_i \leq 1.
\end{equation}

@@

\textbf{Dimensionality reduction.} For solving the optimization problems (\ref{eq:04}) the dimensionality reduction on the base of Peano curves is applied \cite{Gergel2019_2,Gergel2018,GergelKozinov2020,Lera2024,LERA2021125660}. The Peano curve $y(x)$ maps the closed interval $[0,1]$ of the real axis onto the parallelepiped (\ref{eq:02}) unambiguously and continuously. It enables to replace the multidimensional problem (\ref{eq:04}) with solving the equivalent univariate problem
\begin{equation}
\label{eq:07}
\min \left\{\varphi(y), y \in D\right\} = \min\left\{ \varphi(y(x)), x \in [0,1] \right\}.
\end{equation}

@@

As part of the proposed approach, the machine learning model is built according to the rules similar to the construction of the SVC model from section \ref{ssec:41}. As before, all trial points are divided into two classes -- points of the Pareto set and points that do not belong to the Pareto set. The classifier is trained on the selected classes. Note that the classifier models of most machine learning libraries can return not only the class label, but also the probability of relevance to the class. Unlike the approach described in Subsection \ref{ssec:41} the characteristic $R_{ps}(i)$ is computed as the average of probabilities of situations when the points
%$y_i=y(x_i)$ and $y_{i-1}=y(x_{i-1})$
corresponding to boudaries of the interval $(x_{i-1},x_{i})$ belong to the Pareto set.

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.8\textwidth]{fig1.png}
\caption{Heatmap of the function $R_{ps}(i)$ in the criteria space for algorithms with different machine learning models.} 
\label{fig:1}
\end{figure}

@@

First, when changing the optimization problem from set (\ref{eq:07}), a new weight calculation scheme for each of the two classes of trial points is proposed. In this scheme, the weights of each class are calculated inversely proportional to the ratio of the number of all points to the number of points belonging to the Pareto set.

@@

Computational experiments were conducted for each method and each model in order to assess the effectiveness of using these techniques in the global search algorithm. Fig. \ref{fig:2} demonstrates the visualization of $R_{ps}(i)$ similar to Fig. \ref{fig:1}.
\begin{figure}[t!]
\centering
\includegraphics[width=0.8\textwidth]{fig2.png}
\caption{Heatmap of the function $R_{ps}(i)$ in the criteria space for algorithms with different techniques of building models after transformation (\ref{eq:18}).} 
\label{fig:2}
\end{figure}

@@

\textbf{Experiments with fixed weights of two trial classes.} The first series of experiments is aimed at comparison of the algorithm ML\_MGSA\_Prob (Fixed weights) with results of running for methods MGSA and ML\_MGSA\_Dist.

@@

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
% \usepackage{graphicx}
\begin{table}[t]
\centering
\caption{Results of computational experiments}
\label{tab:1}
\resizebox{\textwidth}{!}{%
\begin{tabular}{cccccc}
\hline
\textbf{}                                                                  & \textbf{Model Kernel} & \textbf{$\alpha$} & \textbf{Avg. HV-index} & \textbf{Avg. iter.} & \textbf{Speedup} \\ \hline
ML\_MGSA                                                                   & -          & -    & 114.91  & 2269.82   & 1.00           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of fixed weights}}                                                                \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 114.67  & 691.14    & \textbf{3.28}  \\
\multirow{3}{*}{ML\_MGSA\_Prob (Fixed weights)}                            & Linear     & 0.04 & 114.72  & 963.05    & 2.36           \\
                                                                           & Poly       & 0.07 & 114.73  & 871.82    & 2.60           \\
                                                                           & RBF        & 0.03 & 114.80  & 883.04    & 2.57           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of adjusted weights}}                                                             \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 114.67  & 691.14    & \textbf{3.28}  \\
\multirow{3}{*}{ML\_MGSA\_Prob (Adj. weights)}                             & Linear     & 0.04 & 114.76  & 1018.03    & 2.23           \\
                                                                           & Poly       & 0.08 & 114.73  & 844.66    & 2.69           \\
                                                                           & RBF        & 0.03 & 114.76  & 796.74    & 2.85           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments for the case of adjusted weights and logarithmic probabilities}}                                               \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 114.67  & 691.14    & \textbf{3.28}  \\
\multirow{3}{*}{ML\_MGSA\_Prob (Log Norm)}                                 & Linear     & 0.05 & 114.72  & 711.44    & 3.19           \\
                                                                           & Poly       & 0.04 & 114.79  & 854.82    & 2.66           \\
                                                                           & RBF        & 0.04 & 114.74  & 704.99    & 3.22           \\ \hline
\multicolumn{6}{c}{\textbf{Experiments with 4-criterial MCO problems}}                                                                \\ \hline
ML\_MGSA                                                                   & -          & -    & 7819.86 & 6046.4    & 1.00           \\ \hline
ML\_MGSA\_Dist                                                             & Linear     & 0.01 & 7751.43 & 1925.6    & 3.14           \\
\multirow{3}{*}{ML\_MGSA\_Prob (Log Norm)}                                 & Linear     & 0.08 & 7729.71 & 1760.5    & 3.43           \\
                                                                           & Poly       & 0.08 & 7796.20 & 4121.7    & 1.47           \\
                                                                           & RBF        & 0.08 & 7786.48 & 1364.6    & \textbf{4.43}  \\ \hline
\end{tabular}%
}
\end{table}

Fig. \ref{fig:3} shows the change of the average HV-index as a function of the number of trials.
\begin{figure}[t!]
\centering
\includegraphics[width=0.8\textwidth]{fig3.png}
\caption{Comparison of HV-indices for the case of fixed weights.} 
\label{fig:3}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.8\textwidth]{fig4.png}
\caption{HV-indices comparison  in the case of adjusted weights} 
\label{fig:4}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.8\textwidth]{fig5.png}
\caption{Comparison of  HV-indices for methods with probabilities (\ref{eq:18})} 
\label{fig:5}
\end{figure}

@@

\begin{figure}[t!]
\centering
\includegraphics[width=0.8\textwidth]{Fig6.png}
\caption{Efficiency comparison in the 4-critarial case} 
\label{fig:6}
\end{figure}
